---
title: "black federal workers"
format: html
editor: visual
---

# Setup

```{r, setup = T}
library(tidycensus)
library(tigris)
library(janitor)
library(tidyverse)
library(mapview)
library(survey)
source("functions/setup_functions.R")
```

# Fact-check

Charles and Prince George’s counties, Maryland neighbors with intertwined histories, are the nation’s richest majority-Black counties, with median household incomes that exceed \$100,000. Nowhere else comes close.

```{r, warning = F, echo = F, message = F}

# You will need a tidycensus API Key

# load data on race + fedeal employment + income from census
us_counties_feds <- 
  get_acs(
  # load at the county level
  geography = "county",
  # we want 5-year-ACS from 2023
  survey = "acs5",
  year = 2023,
  variables = c("B24080_009", # fed workers (men)
                "B24080_019", # fed workers (women)
                "B24080_001", # total workers 
                "B19013_001", # median hh income 
                "B02001_001", # total pop (race)
                "B02001_003", # Black or AA alone
                "B02001_002", # white alone
                "B02001_008"
                ),
  output = "wide",
  geometry = T) %>% 
  rename(
    fed_men = B24080_009E,
    fed_women = B24080_019E,
    tot_workers = B24080_001E,
    median_hh_income = B19013_001E,
    total_pop = B02001_001E,
    black_aa = B02001_003E,
    white_pop = B02001_002E,
    two_or_more_races_pop = B02001_008E
  ) %>% 
  mutate(
    # total federal workers
    fed_tot = fed_men + fed_women,
    # federal workers as a percentage of all workers 
    fed_pct = fed_tot / tot_workers * 100,
    # Black population %
    black_pct = black_aa / (total_pop - two_or_more_races_pop) * 100,
    # white population %
    white_pct = white_pop / (total_pop - two_or_more_races_pop) * 100,
    # Two or more population %
    two_or_more_pct = two_or_more_races_pop / total_pop * 100
         ) %>% 
  select(GEOID, 
         NAME, 
         fed_tot,
         tot_workers, 
         fed_pct,
         black_pct,
         white_pct,
         two_or_more_pct,
         median_hh_income) %>% 
    mutate(is_maryland = str_detect(NAME, ", Maryland"),
         is_pg_charles = str_detect(NAME, "Prince George's County, Maryland|Charles County, Maryland")) 


# all majority-Black counties
us_counties_feds %>% 
  filter(black_pct >= 50) %>%
  # ggplot(aes(y = median_hh_income, x = black_pct)) +
  # geom_point(aes(color = is_maryland)) 
  # geom_smooth()
  mapview(zcol = "median_hh_income")

## table 
us_counties_feds %>% 
  filter(black_pct >= 50) %>% 
  st_drop_geometry() %>% 
  as.data.frame() %>% 
  arrange(desc(median_hh_income)) %>% 
  ggplot(aes(y = fed_pct, x = median_hh_income, size = tot_workers, color = is_pg_charles)) +
  geom_point() +
  theme_minimal()




  


```

More than 1 in 8 of workers across the two counties is employed by the federal government. For Black workers, it’s about 1 in 5 – double the statewide average, a Banner analysis of individual-level U.S. Census data shows.

```{r, warning = F, echo = F, message = F}
fed_prop_black_pg_charles <- 
  svymean(~classwkrd_lbl, black_pg_charles_survey) %>% 
  as.data.frame() %>% 
  mutate(category = rownames(.)) %>% 
  tibble() %>% 
  mutate(category = gsub("classwkrd_lbl", "", category)) %>% 
  select(category, everything()) %>% 
  ungroup() %>% 
  mutate(category = str_c(category, " (PG & Charles - Black workers)"))

fed_prop_non_black_pg_charles <- 
  svymean(~classwkrd_lbl, non_black_pg_charles_survey) %>% 
  as.data.frame() %>% 
  mutate(category = rownames(.)) %>% 
  tibble() %>% 
  mutate(category = gsub("classwkrd_lbl", "", category)) %>% 
  select(category, everything()) %>% 
  ungroup() %>% 
  mutate(category = str_c(category, " (PG & Charles - Non-Black workers)")) 

fed_prop_pg_charles <- 
  svymean(~classwkrd_lbl, pg_charles_survey) %>% 
  as.data.frame() %>% 
  mutate(category = rownames(.)) %>% 
  tibble() %>% 
  mutate(category = gsub("classwkrd_lbl", "", category)) %>% 
  select(category, everything()) %>% 
  ungroup() %>% 
  mutate(category = str_c(category, " (PG & Charles)")) 
  

fed_prop_md <- 
  svymean(~classwkrd_lbl, md_survey) %>% 
  as.data.frame() %>% 
  mutate(category = rownames(.)) %>% 
  tibble() %>% 
  mutate(category = gsub("classwkrd_lbl", "", category)) %>% 
  select(category, everything()) %>% 
  ungroup() %>% 
  mutate(category = str_c(category, " (All MD)")) 
  
fed_prop_black_md <- 
  svymean(~classwkrd_lbl, black_md_survey) %>% 
  as.data.frame() %>% 
  mutate(category = rownames(.)) %>% 
  tibble() %>% 
  mutate(category = gsub("classwkrd_lbl", "", category)) %>% 
  select(category, everything()) %>% 
  ungroup() %>% 
  mutate(category = str_c(category, " (All MD - Black workers)")) 

fed_prop_bmore <- 
  svymean(~classwkrd_lbl, bmore_survey) %>% 
  as.data.frame() %>% 
  mutate(category = rownames(.)) %>% 
  tibble() %>% 
  mutate(category = gsub("classwkrd_lbl", "", category)) %>% 
  select(category, everything()) %>% 
  ungroup() %>% 
  mutate(category = str_c(category, " Baltimore + county")) 

fed_prop_non_black_bmore <- 
  svymean(~classwkrd_lbl, non_black_bmore_survey) %>% 
  as.data.frame() %>% 
  mutate(category = rownames(.)) %>% 
  tibble() %>% 
  mutate(category = gsub("classwkrd_lbl", "", category)) %>% 
  select(category, everything()) %>% 
  ungroup() %>% 
  mutate(category = str_c(category, " Baltimore + county - Non-Black workers")) 




fed_prop_black_bmore <- 
  svymean(~classwkrd_lbl, black_bmore_survey) %>% 
  as.data.frame() %>% 
  mutate(category = rownames(.)) %>% 
  tibble() %>% 
  mutate(category = gsub("classwkrd_lbl", "", category)) %>% 
  select(category, everything()) %>% 
  ungroup() %>% 
  mutate(category = str_c(category, " Baltimore + county - Black workers")) 
  

# viz it up
fed_prop_comparison <- 
  fed_prop_black_pg_charles %>% 
  full_join(fed_prop_non_black_pg_charles) %>% 
  full_join(fed_prop_md) %>% 
  full_join(fed_prop_pg_charles) %>% 
  full_join(fed_prop_black_bmore) %>%
  full_join(fed_prop_non_black_bmore) %>% 
  filter(str_detect(category, "Federal govt employee")) %>% 
  arrange(desc(mean))

fed_prop_comparison %>% 
  ggplot(
        aes(x = reorder(category, -mean), y = mean, fill = category)
  ) +
  geom_col() +
  theme_minimal() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()
  ) 

fed_prop_comparison %>% 
  mutate(
    mean = mean * 100,
    SE = SE * 100
         ) %>% 
  select(-SE) %>% 
  kbl() %>% 
  kable_material_dark()


```

Unlike in other communities, federal employment is the main determinant of income status in Maryland’s Black neighborhoods. Here, each 1% jump in workers employed by the federal government adds about \$3,300 in median household income.

```{r, warning = F, echo = F, message = F}
maj_black_tracts %>% 
  mutate(fed_pct = fed_pct / 100) %>% 
  ggplot((aes(x = median_hh_income, y = fed_pct))) +
  geom_point(color = "#267DFF") +
  geom_smooth(method = "lm", color = "#FF4019") +
  labs(
    title = "% of Federal workers vs. Median household income in Majority-Black Census Tracts",
    subtitle = "Every 1% of federal workers in a tract explains about $3300 in household income"
  ) +
  xlab("Median household income") +
  ylab("% federal workers") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(labels = scales::dollar) +
  theme_minimal()

# regression 
maj_black_linear_regression <- lm(median_hh_income ~ fed_pct, data = maj_black_tracts) %>%
  summary()

maj_black_linear_regression

```

When Franklin D. Roosevelt integrated the federal workforce through executive order in 1941, Prince George’s County was over 80% white and mostly rural.

```{r}
census_design <- svydesign(
  ids = ~0,
  weights = ~PERWT,   # Person weight variable
  data = pg_micro_data_clean_1940
)


# Create a weighted contingency table of race by county
race_table <- svytable(~race_lbl, design = census_design)

race_table %>% 
  as.data.frame() %>% 
  kbl() %>% 
  kable_material_dark()

# Convert to percentages (within each county)
race_percentages <- prop.table(race_table) * 100

race_percentages %>% 
  as.data.frame() %>% 
  kbl() %>% 
  kable_material_dark()


```

Today, over 70% of full-time Black workers in Prince George’s County are homeowners, nearly twice the rate of Black Americans as a whole.

```{r}
black_pg_charles_survey_homeownership <- svy_design_function(pg_charles_analysis %>% filter(OWNERSHP != 0) %>% filter(UHRSWORK >= 40))


homeownership_by_fedstatus <-
  svymean(
  ~I(ownershp_lbl),  # Owned home indicator
  # ~I(is_federal),  # Federal worker indicator
  design = black_pg_charles_survey_homeownership,
  FUN = svymean,
  na.rm = TRUE
  ) %>% 
  as.data.frame() %>% 
  mutate(ownership = rownames(.)) %>% 
  select(ownership, everything()) %>% 
  clean_names() %>% 
  mutate(
    ownership = str_remove(ownership, "^I\\(ownershp_lbl\\)")
  ) 

rownames(homeownership_by_fedstatus) <- 1:nrow(homeownership_by_fedstatus)



homeownership_by_fedstatus %>% 
  kbl() %>% 
  kable_material_dark()


```

Federal employment has also been a gateway into the middle class for Black workers without college degrees, who earn median incomes more than \$30,000 higher than other Black non-college educated workers.

```{r}
black_pg_charles_survey_full_time_non_college <- svy_design_function(black_pg_charles_analysis %>% filter(UHRSWORK >= 40) %>% filter(EDUC < 10))

black_pg_charles_inc_quantile_fedstatus_non_college <- svyby(
  ~I(INCWAGE),  # Owned home indicator
  ~I(is_federal),  # Federal worker indicator
  design = black_pg_charles_survey_full_time_non_college ,
  FUN = svyquantile,                      
  quantiles = c(0.25, 0.5, 0.75),
  na.rm = TRUE
  ) %>% 
  rename_with(
    ~gsub("I\\(INCWAGE\\)\\.", "", .), 
    starts_with("I(INCWAGE)")
  ) %>% 
  select(-starts_with("se")) %>% 
  tibble() %>% 
  rename(is_federal = "I(is_federal)")

black_pg_charles_inc_quantile_fedstatus_non_college %>% 
  kbl() %>% 
  kable_material_dark()

black_pg_charles_inc_quantile_fedstatus_non_college %>% 
    pivot_longer(
    cols = c("0.25", "0.5", "0.75"),
    names_to = "quantile",
    values_to = "income"
  ) %>% 
  ggplot(aes(x = quantile, y = income, fill = quantile)) +
  geom_col()+
  facet_wrap(~is_federal)
```
